---
title: "HW2 Kathleen Nie"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part A

```{r warning=FALSE, include=FALSE}

library(DataExplorer)
library(tidyverse)# includes dplyr, ggplot2 and other pkgs
library(lubridate)# for working with dates
library(spData)
library(ISLR)

library(maps)     
library(rnaturalearth)
```

```{r}

col = c("character", 
    "integer", 
    "integer",   
    "character",  
    "character",   
    "character",
    "character",
    "character",
    "real",
    "real",
    "integer",
    "integer",
    "integer",
    "character",
    "integer",
    "integer",
    rep("NULL", 147))

cols = c(rep("integer", 6), "double", "real", rep("character", 8), rep("NULL", 147))

ourdata <- read.table(file = "C:/Users/kathl/Downloads/ibtracs.NA.list.v04r00.csv", sep = ",", colClasses = col, skip = 2, na.strings = c(" ", "hurdat_atl", "atcf", "hurdat_epa"))
```

```{r}
colnames(ourdata) <- c("SID", "SEASON", "NUMBER", "BASIN", "SUBBASIN", "NAME", "ISO_TIME", "NATURE", "LAT", "LON", "WMO_WIND", "WMO_PRES", "WMO_AGENCY", "TRACK_TYPE", "DIST2LAND", "LANDFALL")


```
# Part B
```{r}
summary(ourdata)
```

```{r}
correlation = subset(ourdata, select = -c(SID, NAME, ISO_TIME, WMO_WIND, WMO_PRES, LANDFALL) )
plot_correlation(correlation, title = "Cross Correlation Plot")

```

# Part C

```{r}

data2012 = subset(ourdata, SEASON == 2012)

```
## C.1) 


a) there were a total number of 19 depressions 
```{r}

summarise(group_by(filter(data2012, WMO_WIND < 33), SID))


```

b) there were a total of 19 named storms 
```{r}
nrow(summarise(group_by(data2012, NAME)))
```

c) 10 named hurricanes with their names displayed below

```{r}
summarise(group_by(filter(data2012, WMO_WIND > 64), NAME))
```

d) 2 major hurricanes with their names displayed below 

```{r}
summarise(group_by(filter(data2012, WMO_WIND > 96), NAME))
```

e) first system formed was Alberto on 5/19/2012
```{r}
temp <- arrange(filter(data2012, SEASON == 2012), ISO_TIME)
slice(temp,1)
```



f) Last system dissipated was SANDY on 10/31/2012

```{r}
slice_tail(temp, n=1)
```

## C.2) 

a) 
```{r}
head(select(arrange(data2012, desc(WMO_WIND)), NAME) , n=1)
```
b) 
```{r}
head(select(arrange(data2012, desc(WMO_WIND)), WMO_WIND) , n=1)
```
c) 

```{r}
head(arrange(select(subset(data2012, NAME == "TONY"), WMO_PRES), WMO_PRES), n = 1)
```

# C.3) 

a) the historical average number of storms is about 6 per year. In 2012, we had 19 as calculated in C.1). The 2012 season is above-average
```{r}
nrow(summarise(group_by(ourdata, NAME)))/(2021-1970)
```


b)
the historical average number of hurricanes is about 5. From C.1) we saw 2012 had 10 hurricanes. 2012 season was above-average
```{r}
nrow(summarise(group_by(filter(ourdata, WMO_WIND > 64), NAME))) /(2021-1970)

```


c) from C.2) we got our max wind speed as 100 in 2012. The historical average wind speed shown below is 115. The 2012 season was below average for wind speed

```{r}
maxes = summarise(group_by(filter(ourdata, SEASON < 2021), SEASON), max = max(WMO_WIND, na.rm = TRUE))
mean(maxes$max)
```

## C.4) 

a) Season Summary Map 
```{r}
world_map <- map_data("world")

gg_world <- ggplot() + 
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group), 
               fill = "gray95", colour = "gray70", size = 0.2) +
  theme_bw() + ggtitle("Season Summary Map")

gg_world +
  geom_point(data = data2012,
             aes(x = LON, y = LAT, color = NAME)) +
xlim(c(-130, 0)) + 
  ylim(c(0, 70))
```




b) 
yes, 5 (if I didn't miscount)
they're Isaac, Beryl, Sandy, Alberto, and Debby (if I'm not color blind)

