---
title: "Lab 9: Loops"
subtitle: "Stat 133, Fall 2022"
output: 
  html_document:
    toc: true
---

> ### Learning Objectives
>
> - Get familiar with the syntax of a `for` loop
> - Get familiar with the syntax of a `while` loop


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### General Instructions

- Write your descriptions, explanations, and code in an `Rmd` (R markdown) file.
- Name this file as `lab09-first-last.Rmd`, where `first` and `last` are your
first and last names (e.g. `lab09-gaston-sanchez.Rmd`).
- Knit your `Rmd` file as an html document (default option).
- Submit your `Rmd` and `html` files to bCourses, in the corresponding lab assignment.


-----


## About Loops

- Many times we need to perform a procedure several times
- In other words, we have to perform the same operation several times as long as some condition is fulfilled
- For this purpose we use loops
- The main idea is that of __iteration__ or __repetition__
- R provides three basic paradigms to handle this situations: `for`, `repeat`, `while`


-----


## 1) Counting Letters

Consider the following vector `letrs` which contains various letters:

```{r}
letrs <- c(
  'y', 'd', 'g', 'a', 'b', 'w', 'k', 'n', 'r', 's', 
  'a', 'u', 'u', 'j', 'v', 'n', 'j', 'g', 'i', 'o', 
  'u', 'e', 'i', 'y', 'n', 'e', 'e', 'b', 'j', 'y', 
  'l', 'o', 'a', 't', 'c', 'f', 'j', 'j', 'f', 'o', 
  't', 't', 'z', 'l', 'y', 'w', 'f', 'y', 'h', 'l', 
  'y', 'w', 'x', 'f', 'z', 'g', 's', 'j', 'f', 'x', 
  'n', 'b', 'm', 'r', 'v', 'n', 'f', 'a', 's', 's', 
  'h', 'f', 'w', 'l', 'f', 'h', 'g', 'k', 'q', 'd', 
  'm', 'h', 'y', 'p', 'y', 'w', 'n', 't', 'g', 'm', 
  'v', 'l', 'p', 'a', 'm', 'u', 'f', 'q', 'i', 'g'
)
```


### 1.1) Counting `a`'s with vectorized code

Say we are interested in counting the number of letters `"a"`. This can be 
easily done in R with some vectorized code:

```{r}
sum(letrs == 'a')
```

For learning purposes, we are going to ask you to forget about vectorization
for a moment. And instead let's see how to use __loops__.


### 1.2) Counting `a`'s with a `for` loop

Alternatively, we can also write a `for` loop that iterates through each 
element of `letrs`, testing whether we have an `"a"`, and if yes, the count 
increases by one.

```{r}
# start at count zero
count_a = 0

for (pos in 1:length(letrs)) {
  # increase count if letter is an 'a'
  if (letrs[pos] == 'a') {
    count_a = count_a + 1
  }
}

count_a
```


### 1.3) Your Turn: number of letters different from `"a"`

Consider the vector `letrs`. Adapt the __for__ loop in order to count
the number of letters __different__ from `"a"`

```{r}
# your code
count_not = 0
for (pos in 1:length(letrs)) {
  # increase count if letter is an 'a'
  if (letrs[pos] != 'a') {
    count_not = count_not + 1
  }
}

count_not


```


-----


## 2) Counting `a`'s with a `while()` loop

Alternatively, we can also write a `while` loop that iterates through each 
element of `letrs`, testing whether we have an `"a"`, and if yes, the count 
increases by one.

```{r}
# start at count zero
count_a = 0

# initialize iterator
pos = 1


while (pos <= length(letrs)) {
  # increase count if letter is an 'a'
  if (letrs[pos] == 'a') {
    count_a = count_a + 1
  }
  # increase iterator
  pos = pos + 1
}

count_a
```


### 2.1) Your Turn: number of `a` or `u`

Consider the vector `letrs`. Adapt the __while__ loop in order to count
the number of letters equal to `a` or `u`

```{r}
# your code
count = 0

# initialize iterator
pos = 1


while (pos <= length(letrs)) {
  # increase count if letter is an 'a'
  if (letrs[pos] == 'a') {
    count = count + 1
  } 
  if (letrs[pos] == 'u') {
    count = count + 1
  }
  # increase iterator
  pos = pos + 1
}

count

```


-----


## 3) Your Turn: number of `f` or `w` in even positions

Consider the vector `letrs`. Let's focus on the number of letters equal to `f` 
or `w` that are in even positions (e.g. 2, 4, ..., 100)


### 3.1) Your Turn: `for` loop

Write a __for__ loop in order to count
the number of letters equal to `f` or `w` that are in even positions (e.g. 
2, 4, ..., 100). _Hint_ the function `seq()` is your friend.

```{r}
# your for loop
count_3 = 0
for (pos in 1:(length(letrs)/2)) {
  # increase count if letter is an 'a'
  if (letrs[pos*2] == 'f') {
    count_3 = count_3 + 1
  }
  if (letrs[pos*2] == 'w') {
    count_3 = count_3 + 1
  }
}

count_3

```


### 3.2) Your Turn: `while` loop

Write a __while__ loop in order to count
the number of letters equal to `f` or `w` that are in even positions (e.g. 
2, 4, ..., 100).

```{r}
# your while loop
count_32 = 0

# initialize iterator
pos = 1


while (pos <= (length(letrs)/2)) {
  # increase count if letter is an 'a'
  if (letrs[pos*2] == 'f') {
    count_32 = count_32 + 1
  } 
  if (letrs[pos*2] == 'w') {
    count_32 = count_32 + 1
  }
  # increase iterator
  pos = pos + 1
}

count_32

```


-----


## 4) Stopping a loop with `break`

Say we are interested in counting the number of `x`, `y` and `z` letters,
using a while loop, but this time, we only want to count until we get the 
fifth occurrence. When this happens, we `break` the loop.

```{r}
# start at count zero
count_xyz = 0

for (pos in 1:length(letrs)) {
  # increase count if letter is x, y, or z
  if (letrs[pos] %in% c('x', 'y', 'z')) {
    count_xyz = count_xyz + 1
  }
  # break loop if count gets to fifth occurrence
  if (count_xyz == 5) {
    break
  }
}

count_xyz
```



### 4.1) Your Turn: counting vowels

Consider the vector `letrs`. Write a __for__ loop in order to count
the number of vowels, until reaching exactly 15 vowels. How many iterations were
necessary to obtain 15 vowels?

```{r}
# your code

count_v = 0
iter = 0

for (pos in 1:length(letrs)) {
  # increase count if letter is x, y, or z
  iter = iter + 1
  if (letrs[pos] %in% c('a', 'e', 'i', 'o', 'u')) {
    count_v = count_v + 1
  }
  # break loop if count gets to fifth occurrence
  if (count_v == 15) {
    break
  }
}

iter


```
it took 68 iterations 

-----


## 5) Your Turn: Summation Series

Write a for loop to compute the following series. Your loop should start at 
step $k=0$ and stop at step $n$. Test your code with
different values for $n$. And use a vector to store each k-th term at each 
iteration.
Does the series converge as $n$ increase?

$$
\sum_{k=0}^{n} \frac{1}{2^k} = 1 + \frac{1}{2} + \frac{1}{4} + \frac{1}{8} + \dots + \frac{1}{2^n}
$$

```{r}
# your code
n <- 10
store <- 0

for (i in 0:n) {  
  term <- 1 / (2^i)  
  store[i+1] <- term
}

store
sum(store)

```
yes, the series does converge as n goes toward infinity 

-----


## 6) Your Turn: Geometric Sequence

A sequence such as $3, 6, 12, 24, 48$ is an example of a geometric sequence. 
In this type of sequence, the $n$-th term is obtained as: 

$$
a_n = a_1 \times r^{n-1}
$$

where: $a_1$ is the first term, $r$ is the common ratio, and $n$ is the number of 
terms.

Write a for loop to compute the sum of the first $n$ terms of:
3 + 6 + 12 + 24 + ... 
Test your code with different values for $n$. Does the series converge as $n$ 
increase?

```{r}
# your code
a1 <- 5
r <- 2
sum_geo <- 0
for (n in 1:10) {  
  an = a1 * r^(n-1)  
  sum_geo = sum_geo + an  
}

sum_geo

```
No, the series does not converge as n increases towards infinity 
