---
title: "Stat 151A Hw 6"
author: "Kathleen Nie"
date: "4/30/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 4) 
```{r}
library("ISLR")
library(magrittr)
library(dplyr)
library(MASS)
```

```{r}


data <- Smarket[-c(1, 7,8,9)]
variables = as.matrix(data) 
direction = Smarket$Direction



a = direction == "Up"
y = sapply(a, as.numeric)
y = as.matrix(y)

  
```

```{r echo=TRUE}

compute_p <- function(x,beta){
  X <- cbind(1,x)
  p <- exp(X%*%beta)/(1+exp(X%*%beta)) 
  return(p)
}

compute_var <- function(p){
  W <- p*(1-p) 
  return(diag(as.vector(W)))
}

compute_var_vector <- function(p) {
  W <- p*(1-p)
  return(W)
}

compute_expected_resp <- function(x,beta,y,p,W){
  X <- cbind(1,x)
  z <- X%*%beta+(y-p)/W 
  return(z)
}

b_new <- function(X, W, z) {
  lhs <- t(X)%*%W%*%X
  rhs <- t(X)%*%W%*%z
  return(solve(lhs)%*%rhs)
}

did_we_converge <- function(beta0,beta1,conv.eps){
  sum((beta0-beta1)^2)<=conv.eps
}




IRLS <- function(X,y,max.iter=100,conv.eps=1e-12){
  beta <- rep(0,ncol(cbind(1,X))) 
  beta_prev <- beta     
  

  
  for(iter in 1:max.iter){
    p <- compute_p(X,beta) 
    w <- compute_var(p) 
    w2 <- compute_var_vector(p)
    z <- compute_expected_resp(X,beta,y,p,w2)
    
    
    beta <- b_new(X,w,z)
    
    if(sum((beta-beta_prev)^2)<=conv.eps){
      break
    } else {
      beta_prev <- beta
    }
  }
  
  return(beta)
}
```

## 5) 


```{r}
IRLS <- function(X,y,max.iter=100,conv.eps=1e-12){
  beta <- rep(0,ncol(cbind(1,X))) 
  beta_prev <- beta               #
  
  for(iter in 1:max.iter){
    p <- compute_p(X,beta) 
    
    x <- cbind(1,X)
    
    x_bar = x%*%t(p)
    
    beta = beta_prev + (t(x)%*%x_bar)^-1 %*% (t(x)%*%(y-p))
  
    
    if(sum((beta-beta_prev)^2)<=conv.eps){
      break
    } else {
      beta_prev <- beta
    }
  }
  
  return(beta)
}
  
```




## 6)

```{r}
library(MASS)

summary(birthwt)

birthwt2 <- birthwt%>%
  rename(hypertension = ht)%>%
  mutate(race = ifelse(race == 1, "white", "non-white"))
head(birthwt2)

m0 <- glm(low ~ race, family = binomial, data = birthwt2)
summary(m0)

```
first looking at whether a child born below normal birth weight is affected by race. From results, being white had negative effect on whether the child was below normal birth weight (-0.6954) compared to other races. Looking at the p0value, the results are statistically significant



```{r}
m1 <- glm(low ~ race + smoke, family = binomial, data = birthwt2)
summary(m1)
```
In this model, the variable smoke was taken into consideration. Mothers who smoked during pregnancy had a higher chance (1.1130) of giving birth to a baby that would be below normal birth weight of 2.5 kg. the p-value shows that this result is significant

```{r}
m2 <- glm(low ~ race + smoke + hypertension, family = binomial, data = birthwt2)
summary(m2)
```
 In this model we see that hypertension does increase the likelihood that a child's birth weight will be below 2.5 kg (1.175). This indicator however is not statistically significant with a p-value of >.05.

```{r}
m3 <- glm(low ~ race + smoke * hypertension, family = binomial, data = birthwt2)
summary(m3)
```
lastly, we examine the interaction between smoking and hypertention together. The model shows that people of white race are more likely to have a child that is of normal birth weight or higher. It also shows that smoking have a negative effect. There is however no correlation between hypertension and lower birth weight. =



```{r}
anova(m0, m1, m2, m3, test = "Chisq")
```

Model 4 seems to have the best fit 
